Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-04 text/ms.tex
Tidy up some TODOs
Black swan-> black-swan when an adjective

%TODO look up doi or volume for Drake 2014 Ecology

%\noindent
%\emph{Main message, to be deleted (25 words)}: Black swans are present \& taxonomically widespread but rare in population dynamics. Extreme climate, predation, parasites, and their interactions are common causes; intrinsic drivers are unclear.

\section{Abstract}

@@

There is compelling anecdotal evidence for ecological black swans, but systematic evidence has been elusive. A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and a number of major environmental events on geologic and contemporary time scales could be considered to have caused ecological black swans \citep{nunez2012}. Black swans, or heavy tailed distributions, have been discovered in time-series of marine microbe populations \citet{segura2013} and time to extinction for experimental waterflea populations \citet{drake2014}. As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which found little clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.
% TODO: pull anecdote from that Nunez study

@@

Here we develop a framework for identifying ecological black swans in the process noise of population dynamics, i.e.\ the stochastic jumps in abundance from one time step to the next. Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and can be easily applied to abundance time series. We apply our model to \NPops\ of populations from four taxonomic classes. We estimate the frequency and strength of black swans in population dynamics across taxonomy; verify some causes of black swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
%We find a high probability of ecological
%black swans across a wide range of taxa but in a limited number of populations.
%When they occur, extreme climate, predation, parasites, and their interactions
%are common documented causes. We find little evidence that intrinsic
%life-history characteristics are related to the probability of black swans.

@@

By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process errors have heavy tails and are thereof evidence of black swan events (Fig.~\ref{fig:didactic}a, b). For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $0.018$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$) TODO CHECK. As the value of $\nu$ approaches infinity, the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).

@@

Across populations, the probability of observing black swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series length was the strongest covariate of observing black swan dynamics. For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with increased time series length was small ($0.\pHeavyNSixty$ vs.\ $0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

% "Perhaps then you could say "we examined all time series with published explanations of why the black swan events occurred"
TODO STATE HOW THESE 7 WERE CHOSEN
TODO QUANTIFY UPS AND DOWNS --- SKEW
We assembled the documented causes of black swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table 1). The majority of documented events were downward black swans and involved a series of unfortunate events. For example, a shortage of nest sites for the shag long-necked bird (TODO SCI.\ NAME) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black swan for a water vole population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings in grey heron abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.

@@

We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series. Black swans are detected more frequently in longer time series and in populations with a smaller magnitude of process error. However, black swans were not associated with density dependence, population growth rate, or lifespan. In verified cases, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black swan population dynamics and if anything will underestimate the probability of black swans. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.

Our results clarify previous related analyses. Previous work with an older version of the GPDD assessed the distribution of abundance time series but focused on identifying if the log-normal distribution was the most frequently parsimonious model \citep{halley2002}. For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series. However, black swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes. Therefore, we would not expect to observe black swans in a large proportion of populations. By probabilistically quantifying process noise tails on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans. In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance. However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could falsely generate heavy tails. This rebuttal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with the same mean and inverse-gamma-distributed variances,][]{gelman2014}.

We might expect to observe black swan dynamics in ecological time series because of either unmodelled intrinsic properties of populations or extrinsic forces acting on populations. Since a t-distribution can be formed by a mixture of normal distributions \citep{gelman2014}, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}. That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time. Extrinsic forces could also cause black swan dynamics \citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans. Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black swan ecological dynamics. This could occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~1). In addition to a possibly biased sample of populations, some black swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black swan dynamics.

@@

Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by black swan events.
TODO END ON THIS

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions. Ecological resource management can draw from other disciplines that focus on heavy tails. For example, earthquake preparedness and response is focussed on black swan events. Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans (e.g.~REFERENCE?). A similar focus might benefit resource management once our ability to predict the spatial probability and covariates of ecological black swans improves. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For example, setting target population abundances that are appropriately set back from critical limits may buffer black swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}. In recent decades ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes. Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed black swan dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations. Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order and sorted by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer populations in panel a are omitted for space. Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed black swan dynamics.
}
\label{fig:nu-coefs}
\end{center}
\end{figure}
\clearpage

>>>>

% TODO: hyphenate the adjective 'black swan'? E.g. black-swan event, black swans.
\section{Abstract}

@@

There is compelling anecdotal evidence for ecological black swans, but systematic evidence has been elusive. A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and a number of major environmental events on geologic and contemporary time scales could be considered to have caused ecological black swans \citep{nunez2012}. Black swans, or heavy-tailed distributions, have been discovered in time-series of marine microbe populations \citet{segura2013} and time to extinction for experimental waterflea populations \citet{drake2014}. As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which found little clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.
% TODO: pull anecdote from that Nunez study

@@

Here we develop a framework for identifying ecological black swans in the process noise of population dynamics, i.e.\ the stochastic jumps in abundance from one time step to the next. Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and can be easily applied to abundance time series. We apply our model to \NPops\ of populations from four taxonomic classes. We estimate the frequency and strength of black swans in population dynamics across taxonomy; verify some causes of black-swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
%We find a high probability of ecological
%black swans across a wide range of taxa but in a limited number of populations.
%When they occur, extreme climate, predation, parasites, and their interactions
%are common documented causes. We find little evidence that intrinsic
%life-history characteristics are related to the probability of black swans.

@@

By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process errors have heavy tails and are thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b). For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $0.018$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$) TODO CHECK. As the value of $\nu$ approaches infinity, the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).

@@

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series length was the strongest covariate of observing black-swan dynamics. For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with increased time series length was small ($0.\pHeavyNSixty$ vs.\ $0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

% "Perhaps then you could say "we examined all time series with published explanations of why the black swan events occurred"
TODO STATE HOW THESE 7 WERE CHOSEN
TODO QUANTIFY UPS AND DOWNS --- SKEW
We assembled the documented causes of black-swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table 1). The majority of documented events were downward black swans and involved a series of unfortunate events. For example, a shortage of nest sites for the shag long-necked bird (TODO SCI.\ NAME) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black swan for a water vole population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.

@@

We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series. Black swans are detected more frequently in longer time series and in populations with a smaller magnitude of process error. However, black swans were not associated with density dependence, population growth rate, or lifespan. In verified cases, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black swans. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.

Our results clarify previous related analyses. Previous work with an older version of the GPDD assessed the distribution of abundance time series but focused on identifying if the log-normal distribution was the most frequently parsimonious model \citep{halley2002}. For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series. However, black-swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes. Therefore, we would not expect to observe black swans in a large proportion of populations. By probabilistically quantifying process noise tails on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans. In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance. However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could falsely generate heavy tails. This rebuttal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with the same mean and inverse-gamma-distributed variances,][]{gelman2014}.

We might expect to observe black-swan dynamics in ecological time series because of either unmodelled intrinsic properties of populations or extrinsic forces acting on populations. Since a t-distribution can be formed by a mixture of normal distributions \citep{gelman2014}, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}. That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time. Extrinsic forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans. Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black-swan ecological dynamics. This could occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~1). In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

@@

Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by black-swan events.
TODO END ON THIS

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions. Ecological resource management can draw from other disciplines that focus on heavy tails. For example, earthquake preparedness and response is focussed on black-swan events. Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans (e.g.~REFERENCE?). A similar focus might benefit resource management once our ability to predict the spatial probability and covariates of ecological black swans improves. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For example, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}. In recent decades ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes. Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed black-swan dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations. Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order and sorted by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer populations in panel a are omitted for space. Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed black-swan dynamics.
}
\label{fig:nu-coefs}
\end{center}
\end{figure}
\clearpage

### 2014-11-04 text/supp.tex
Black swan-> black-swan when an adjective

\noindent
We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated also only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}). Note that interpolation is highly unlikely to lead to black swan detections, since black swans involve extreme increases or decreases. Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.

@@

We log transformed $\sigma$, time-series length, and lifespan to match the way
they are visually represented in Fig.~\ref{fig:correlates} and to make the
relationship approximately linear on the logit-transformed response scale. All
input variables were standardized by subtracting their mean and dividing by two
standard deviations to make their coefficients comparable in magnitude
\citep{gelman2008c}. We excluded body length as a covariate because it was
highly correlated with lifespan, and lifespan exhibited more overlap across
taxonomy than body length. Lifespan is also more directly related to time and
potential mechanisms driving black swan dynamics.

>>>>

\noindent
We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated also only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}). Note that interpolation is highly unlikely to lead to black-swan detections, since black swans involve extreme increases or decreases. Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.

@@

We log transformed $\sigma$, time-series length, and lifespan to match the way
they are visually represented in Fig.~\ref{fig:correlates} and to make the
relationship approximately linear on the logit-transformed response scale. All
input variables were standardized by subtracting their mean and dividing by two
standard deviations to make their coefficients comparable in magnitude
\citep{gelman2008c}. We excluded body length as a covariate because it was
highly correlated with lifespan, and lifespan exhibited more overlap across
taxonomy than body length. Lifespan is also more directly related to time and
potential mechanisms driving black-swan dynamics.
